---
title: "Bikeshare/Weather Model"
output: html_notebook
author: "Tyler Yankee"
date: "2022-12-07"
---

This analysis was performed with help from the tutorial from Data Science in a Box found here:
https://minecr.shinyapps.io/dsbox-06-dcbikeshare/.

## Load packages
```{r}
library(tidyverse)
library(tidymodels)
library(dsbox)
```


## Clean data

* Recode the `season` variable from numeric values to names, and order them based on the year
* Recode the `holiday` and `workingday` variables from 0/1 to "no/yes"
* Recode the `yr` variable as the absolute year rather than number of years since 2011
* Recode the `weathersit` variable to a description of the weather rather than a number

```{r}
dcbikeshare <- dcbikeshare %>% mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"),
    season = fct_relevel(season, "spring", "summer", "fall", "winter"))

dcbikeshare <- dcbikeshare %>% mutate(
    holiday = ifelse(holiday == 0, "no", "yes"),
    holiday = fct_relevel(holiday, "no", "yes"),
    workingday = ifelse(workingday == 0, "no", "yes"),
    workingday = fct_relevel(workingday, "no", "yes"))

dcbikeshare <- dcbikeshare %>% mutate(
    yr = ifelse(yr == 0, "2011", "2012"),
    yr = fct_relevel(yr, "2011", "2012"))

dcbikeshare <- dcbikeshare %>% mutate(
    weathersit = case_when(
    weathersit == 1 ~ "clear",
    weathersit == 2 ~ "mist",
    weathersit == 3 ~ "light precipitation",
    weathersit == 4 ~ "heavy precipitation"),
    weathersit = fct_relevel(weathersit, "clear", "mist", 
                             "light precipitation", "heavy precipitation"))
```

## New variables

* Create raw values for temperature, humidity, and windspeed
* Check that the sum of `casual` and `registered` adds up to the total `cnt`

```{r}
dcbikeshare <- dcbikeshare %>% mutate(
    temperature_raw = temp * 41,
    feeling_temperature_raw = atemp * 50,
    humidity_raw = hum * 100,
    windspeed_raw = windspeed * 67)

dcbikeshare %>% mutate(cas_plus_reg = casual + registered) %>%
  summarise(all_zero = all(cas_plus_reg == cnt))
```

## Visualizations

```{r}
dcbikeshare %>% ggplot(aes(x = dteday, y = cnt, color = feeling_temperature_raw)) +
    geom_point(alpha = .7) +
    labs(title = "Bike rentals in DC, 2011 and 2012",
         subtitle = "Warmer temperatures associated with more bike rentals",
         x = "Date", y = "Bike rentals", color = "Temperature (C)") +
  theme_minimal()
```

## Fitting a model to the data

```{r}
cnt_tmp <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ temperature_raw, data = dcbikeshare)
cnt_tmp %>%
  tidy()
glance(cnt_tmp)$r.squared
```

A couple takeaways from the model:

* Since the slope parameter (for `temperature_raw`) is about 162, this means that for each additional degree Celsius, we can expect bike rentals to increase by about 162.
* The $R^2$ value of .394 indicates that 39.4% of the variability in the total number of daily bike rentals is accounted for by this model.

## New model

The model between `temperature_raw` and `cnt` wasn't a very good fit - let's take a look at `feeling_temperature_raw` (the feel-like temperature).

```{r}
cnt_atmp <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ feeling_temperature_raw, data = dcbikeshare)
cnt_atmp %>%
  tidy()
glance(cnt_atmp)$r.squared
```

With an $R^2$ value of .398, this model is *slightly* better at capturing the data than the previous model, but not by much.

## Full model

Let's include all variables impacting the total bike count: `season`, `yr`, `holiday`, `workingday`, `weathersit`, `temperature_raw`, `feeling_temperature_raw`, `humidity_raw`, `windspeed_raw`, and the interaction between `feeling_temperature_raw` and `holiday`.

```{r}
cnt_full <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
tidy(cnt_full)
glance(cnt_full)$adj.r.squared
```

Observations:

* The $R^2$ is much higher for this model, which is a good result.
* Interesting notes on holidays: the intercept for non-holidays is higher than that for holidays, while the slope for feeling_temperature is steeper for holidays than for non-holidays.
* Based on the slope parameters, the season with the highest number of bike rentals appears to be the fall, while the season with the lowest is winter.

## Backward selection

Now that we've made a model with all of the explanatory variables, let's explore removing each of them one by one to see if that provides a better fit.

Without `season`
```{r}
rm_season <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ yr + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_season)$adj.r.squared
```

Without `year`
```{r}
rm_year <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_year)$adj.r.squared
```

Without `holiday` (and interaction between `holiday` and `feeling_temperature_raw`)
```{r}
rm_holiday <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw, data = dcbikeshare)
glance(rm_holiday)$adj.r.squared
```


Without `weathersit`
```{r}
rm_weathersit<- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_weathersit)$adj.r.squared
```

Without `temperature_raw`
```{r}
rm_temp_raw<- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  feeling_temperature_raw + humidity_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_temp_raw)$adj.r.squared
```

Without `feeling_temperature_raw` (and interaction between `feeling_temperature_raw` and `holiday`)
```{r}
rm_feeltemp_raw <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  temperature_raw + humidity_raw +
  windspeed_raw, data = dcbikeshare)
glance(rm_feeltemp_raw)$adj.r.squared
```

Without `humidity_raw`
```{r}
rm_humidity_raw <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw +
  windspeed_raw + feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_humidity_raw)$adj.r.squared
```

Without `windspeed_raw`
```{r}
rm_windspeed_raw <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  feeling_temperature_raw * holiday, data = dcbikeshare)
glance(rm_windspeed_raw)$adj.r.squared
```


Without interaction between `feeling_temperature_raw` and `holiday`
```{r}
rm_interaction <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cnt ~ season + yr + holiday + workingday + weathersit + 
  temperature_raw + feeling_temperature_raw + humidity_raw +
  windspeed_raw, data = dcbikeshare)
glance(rm_interaction)$adj.r.squared
```

Observations:

* When observing the adjusted $R^2$ of all of the above and comparing that to the $R^2$ of the full model, we can conclude that none of these models provide a significant improvement over the full model.
